---
title: "Introductions are in Order, R you Ready?"
output: html_notebook
---

This notebook contains code from Chapters 1 and 2 of [Prelude to Econometrics Using R](https://bookdown.org/mattdobra/Prelude/).  

The companion site to [Prelude to Econometrics Using R](https://bookdown.org/mattdobra/Prelude/) can be found at [https://mattdobra.github.io/EconometricsLectureNotes/](https://mattdobra.github.io/EconometricsLectureNotes/).

This first code chunk isn't in the book, but is useful for installing most/all of the intensively used packages in the book.  If you want to execute this code, only do it once!

```{r}
install.packages('knitr')
install.packages('tidyverse')
install.packages('kableExtra')
install.packages('AER')
install.packages('stargazer')
install.packages('wooldridge')
install.packages('fivethirtyeight')
install.packages('sandwich')
install.packages('lmtest')
install.packages('margins')
install.packages('MASS')
install.packages('jtools')
install.packages('mlogit')
install.packages('censReg')
install.packages('sampleSelection')
install.packages('scales')
install.packages('dynlm')
install.packages('tseries')
install.packages('forecast')
install.packages('cowplot')
install.packages('tidyquant')
install.packages('plm')
```

Remember, only install things once! After that, just load them with the library command!
Libraries needed for this notebook:

```{r}
library(tidyverse)
library(wooldridge)
library(AER)
```


# Chatper 1

```{r}
data(mtcars)
mean(mtcars$mpg[mtcars$cyl==4]) 
mean(mtcars$mpg[mtcars$cyl==6]) 
test1 <- t.test(mtcars$mpg[mtcars$cyl==6], mtcars$mpg[mtcars$cyl==4]) 
pander::pander(test1, split.cells = c(1,1,50,1,1,1)) # Will not work unless you have pander package installed
```

# Chapter 2

## Assigning Objects and Basic Math

```{r echo = TRUE}
q <- 42
```

```{r echo = TRUE}
print(q)
```

```{r echo = TRUE}
q
```


```{r, error = TRUE,  echo = TRUE}
Q
```


```{r  echo = TRUE}
q.1 <- 2.718
q_1 <- 3.142
```


```{r echo = TRUE}
q <- 420
```

```{r echo = TRUE}
q
```

```{r echo = TRUE}
rm(q)
```


```{r echo = TRUE}
rm(list=ls())
```


```{r echo = TRUE}
z <- 132
y <- 33
```


```{r echo = TRUE}
a <- "Hello"
a
```


```{r echo = TRUE}
b <- "1"
```


```{r echo = TRUE}
class(a)
class(b)
class(y)
```


```{r echo = TRUE}
2+2
24-18
45*8
84/4
2^8
abs(-42)
```


```{r echo = TRUE}
z+y
z-y
z*y
z/y
sqrt(y)
log(z)
exp(y)
```


```{r echo = TRUE}
z+4
y^2
z/y+3
```


```{r,  echo = TRUE, error = TRUE}
z+b
```


```{r echo = TRUE}
z+as.numeric(b)
```


```{r,  echo = TRUE, error = TRUE}
z+as.numeric(a)
```


```{r echo = TRUE}
q <- z/y
q
```

## Vectors


```{r echo = TRUE}
num1 <- c(1,4,9,16,25)
num2 <- c(1,3,6,10,15)
```


```{r echo = TRUE}
countries <- c("USA", "Canada", "Mexico")
```

```{r echo = TRUE}
num1-1
num1*3
sqrt(num1)
```


```{r echo = TRUE}
num1+num2
num1*num2
```


```{r echo = TRUE}
num1[5] 
num2[1]
```


```{r echo = TRUE}
num1[-3]
```

```{r echo = TRUE}
num1[num1>5]
```

## Packages and Libraries

```{r echo = TRUE}
ror6 <- 1 + c(.04, .13, -.03, .11, -.05, .08)
```

```{r echo = TRUE}
EnvStats::geoMean(ror6)
```

```{r echo = TRUE}
library(EnvStats)
geoMean(ror6)
```
## Working with Data

```{r echo = TRUE}
data()
```


```{r echo = TRUE}
data(iris)
```

```{r echo = TRUE}
data(CPS1985)
```


```{r echo = TRUE}
dim(CPS1985)
ncol(CPS1985)
nrow(CPS1985)
```

```{r echo = TRUE}
head(CPS1985)
tail(CPS1985)
```

```{r echo = TRUE}
CPS1985[222:225,]
```


```{r echo = TRUE}
CPS1985[CPS1985$age>60 , ]
```

```{r echo = TRUE}
CPS1985[CPS1985$ethnicity == "cauc" & CPS1985$region == "south" & CPS1985$gender == "female" & CPS1985$married == "no", ]
```


```{r echo = TRUE}
attach(CPS1985)
```



```{r echo = TRUE}
CPS1985[ethnicity == "cauc" & region == "south" & gender == "female" & married == "no", ]
```

  
```{r echo = TRUE}
CPS1985[gender == "male" & occupation != "management" & age>55, -7]
```

```{r echo = TRUE}
str(CPS1985)
```


```{r echo = TRUE}
class(wage)
class(age)
class(married)
class(union)
```

```{r echo = TRUE}
class(CPS1985$union)
```

```{r echo = TRUE}
levels(ethnicity)
levels(occupation)
```

```{r echo = TRUE}
summary(CPS1985)
```


```{r echo = TRUE}
tempdata <- CPS1985
tempdata$female <- ifelse(tempdata$gender == "female", 1, 0)
tempdata[200:209,c(7,12)]
```


```{r echo = TRUE}
class(tempdata$female)
```



```{r echo = TRUE}
summary(tempdata$female)
```


```{r echo = TRUE}
summary(as.factor(tempdata$female))
```


```{r echo = TRUE}
tempdata$female <- factor(tempdata$female, levels = c(0,1), labels = c("male", "female"))
summary(tempdata$female)
```


```{r echo = TRUE}
summary(wage[gender == "male"])
summary(wage[gender == "female"])
```


```{r echo = TRUE}
femanager <- gender == "female" & occupation == "management"
summary(femanager)
```


```{r echo = TRUE}
tempdata <- cbind(CPS1985, femanager)
class(tempdata$femanager)
```


```{r echo = TRUE}
tempdata$femanager <- factor(tempdata$femanager, levels = c(FALSE,TRUE), labels = c("no", "yes"))
class(tempdata$femanager)
summary(tempdata$femanager)
```

## Basics of dplyr and tidyverse

```{r echo = TRUE}
CPS1985 %>% 
  filter(age>60)
```

```{r echo = TRUE}
CPS1985 %>% 
  filter(ethnicity == "cauc") %>% 
  filter(region == "south") %>% 
  filter(gender == "female") %>% 
  filter(married == "no")
```

```{r echo = TRUE}
CPS1985 %>% 
    filter(gender == "male") %>% 
    filter(occupation != "management") %>% 
    filter(age>55) %>% 
    select(-gender)
```

```{r echo = TRUE}
CPS1985 %>% 
    group_by(gender) %>% 
    summarize(wage = mean(wage))
```

