---
title: "Descriptive Statistics"
output: html_notebook
---

This notebook contains code from Chapter 3 of [Prelude to Econometrics Using R](https://bookdown.org/mattdobra/Prelude/).  

The companion site to [Prelude to Econometrics Using R](https://bookdown.org/mattdobra/Prelude/) can be found at [https://mattdobra.github.io/EconometricsLectureNotes/](https://mattdobra.github.io/EconometricsLectureNotes/).

Libraries needed for this notebook:

```{r}
library(tidyverse)
library(wooldridge)
library(AER)
library(stargazer)
```

Preload and attach data:

```{r}
data(CPS1985)
attach(CPS1985)
data(vote1)
attach(vote1)
```

# Chatper 3

## Bar and Pie Charts

```{r echo = TRUE}
CPS1985 %>% ggplot(aes(x = occupation)) +
    geom_bar()
```

```{r echo = TRUE}
jobtype <- table(occupation)
jobtype
```

```{r echo = TRUE}
pie(jobtype)
```

```{r echo = TRUE}
set.seed(8675309)
a <- sample(1:6, 100, replace = TRUE)
b <- table(a)
b
```

```{r echo = TRUE}
pie(b)
```


```{r echo = TRUE}
barplot(b)
```

## Boxplots

```{r echo = TRUE}
boxplot(wage)
```

```{r echo = TRUE}
CPS1985 %>% ggplot(aes(y = wage)) +
    geom_boxplot() 
```

```{r echo = TRUE}
CPS1985 %>% ggplot(aes(y = wage)) +
    geom_boxplot() +
    stat_boxplot(geom = "errorbar", width = 0.5)
```

```{r echo = TRUE}
boxplot(wage ~ occupation)
```

```{r echo = TRUE}
CPS1985 %>% ggplot(aes(x = occupation, y = wage)) +
    geom_boxplot() +
    stat_boxplot(geom = "errorbar", width = 0.5)
    
```

```{r echo = TRUE}
CPS1985 %>% ggplot(aes(x = gender, y = wage)) +
    geom_boxplot() +
    stat_boxplot(geom = "errorbar", width = 0.5)
```


```{r echo = TRUE}
CPS1985 %>% ggplot(aes(x = married, y = wage, fill = gender)) +
    geom_boxplot() 
```

```{r echo = TRUE}
CPS1985 %>% ggplot(aes(x = gender, y = wage, fill = married)) +
    geom_boxplot() 
```

```{r echo = TRUE}
CPS1985 %>% ggplot(aes(x = married, y = wage, fill = gender)) +
    geom_boxplot() +
    theme_classic() +
    labs(title = "Does Marriage Influnce the Gender Wage Gap?", 
         subtitle = "1985 CPS data",
         x = "Marital Status", 
         y = "Wages",
         fill = "",
         caption = "@mattdobra") +
    scale_fill_manual(values= c("deepskyblue4", "darksalmon")) +
    theme(legend.position = "bottom",
          legend.direction = "horizontal",
          axis.ticks.x = element_blank()) +
    scale_x_discrete(labels = c("no" = "Unmarried", "yes" = "Married"))
```

## Histograms

```{r echo = TRUE}
hist(age)
```

```{r, message = FALSE, echo =TRUE}
CPS1985 %>%  ggplot(aes(x = age)) +
    geom_histogram()
```

```{r echo = TRUE}
CPS1985 %>%  ggplot(aes(x = age)) +
    geom_histogram(binwidth = 5, color = "black")
```

```{r echo = TRUE}
CPS1985 %>%  ggplot(aes(x = age)) +
    geom_histogram(binwidth = 5, fill = "#CBB676", color = "#B4985A") +
    theme(panel.background = element_rect(fill = ("#00573C")))
```

```{r echo = TRUE}
CPS1985 %>%  ggplot(aes(x = age)) +
    geom_histogram(binwidth = 5, fill = "#B4985A", color = "#00573C") +
    theme(
        panel.background = element_rect(fill = "#00573C"),
        panel.grid.major = element_line(color = "#CBB676"),
        panel.grid.minor = element_line(color = "#CBB676"))
```

## Stacked and grouped bar charts


```{r echo=TRUE}
table1 <- table(gender, occupation)
table1
```

```{r echo = TRUE}
barplot(table1)
```

```{r echo = TRUE}
barplot(table1, beside = TRUE)
```

```{r echo = TRUE}
CPS1985 %>% ggplot(aes(x = occupation, fill = gender)) +
    geom_bar()
```

```{r echo = TRUE}
CPS1985 %>% ggplot(aes(x = occupation, fill = gender)) +
    geom_bar() +
    scale_fill_manual(values= c("deepskyblue4", "lightpink2")) 
```

```{r echo = TRUE}
CPS1985 %>% ggplot(aes(x = occupation, fill = gender)) +
    geom_bar() +
    scale_fill_viridis_d()
```


```{r echo = TRUE}
CPS1985 %>% ggplot(aes(x = occupation, fill = gender)) +
    geom_bar(position = "dodge") +
    scale_fill_viridis_d()
```

```{r echo = TRUE}
CPS1985 %>% ggplot(aes(x = occupation, fill = gender)) +
    geom_bar(position = "fill") +
    scale_fill_viridis_d() +
    labs(y = "")
```

## Scatter Plots

```{r echo = TRUE}
head(vote1)
```

```{r echo = TRUE}
plot(voteA ~ shareA)
```

```{r echo = TRUE}
vote1 %>% ggplot(aes(x = shareA, y = voteA)) +
    geom_point()
```

```{r, message=FALSE, echo = TRUE}
vote1 %>% ggplot(aes(x = shareA, y = voteA)) +
    geom_point() +
    geom_smooth(method = lm)
```


```{r, message=FALSE, echo = TRUE}
vote1 %>% ggplot(aes(x = shareA, y = voteA, color = as.factor(democA))) +
    geom_point() 
```


```{r, echo = TRUE, message=FALSE}
vote1 %>% ggplot(aes(x = shareA, y = voteA)) +
    geom_point(color = "#00573C", fill = "#B4985A", shape = 21, size = 2.5) +
    facet_wrap(~democA)
```

## Summarizing Data Numerically: Categorical Variables

```{r echo = TRUE}
table(occupation)
```

```{r echo = TRUE}
table(occupation)/534
table(occupation)/nrow(CPS1985)
table(occupation)/length(occupation)
table(occupation)/dim(CPS1985)[1]
```


```{r echo = TRUE}
table(occupation, gender)
```

## Summarizing Data Numerically: Numerical Variables

```{r, echo = TRUE}
mean(wage)
```

```{r, echo = TRUE}
mean(wage, trim = 0.05)
```

```{r, echo = TRUE}
median(wage)
```


```{r, echo = TRUE}
var(wage)
sd(wage)
```

```{r, echo = TRUE}
sd(wage)^2
sqrt(var(wage))
```

```{r, echo = TRUE}
min(wage)
max(wage)
```

```{r, echo = TRUE}
range(wage)
```

```{r, echo = TRUE}
quantile(wage, .25)
```

```{r, echo = TRUE}
quantile(wage, probs = c(.1, .25, .5, .75, .9))
```

```{r, echo = TRUE}
quantile(wage, probs = c(0,1))
```


```{r, echo = TRUE}
cor(wage, education)
```


```{r, echo = TRUE, message = FALSE}
CPS1985 %>% ggplot(aes(x = education, y = wage)) +
    geom_point() +
    geom_smooth(method = lm)
```

```{r, echo = TRUE}
cor(wage, education, method = "spearman")
```

```{r, echo = TRUE}
summary(wage)
```

```{r, echo = TRUE}
stargazer(CPS1985, type = "text")
```

```{r, echo = TRUE, results = 'asis'}
stargazer(CPS1985, type = "html")
```

