---
title: "Intro to R"
output: html_notebook
---

# Overview 

The first few weeks of this course are devoted to learning the basics of R and using R to review the material from the statistics prerequisite course you took prior to this one.  Learning the basics of R must come first. 

# Writing Scripts

Before we turn to actually writing some code, let's start by talking about how to write code.  Writing your code in a script is essential for a variety of reasons:

* Scripting makes your work **reproducible**.
* Scripting allows you to **document** your code.
* Scripting makes it easier to work on a project over multiple sessions.

A script is essentially a text file that contains a list of commands that R will execute in order.  Rather than type one command at a time in the console window (bottom left of R Studio), you type the series of commands in the script window (top left of R studio) and can run all or parts of the script at once.  

## Scripting best practices

### Comment your work!

You may work on a script with another person who has no idea what your code is all about, or you may not look at a script for a few weeks and have forgotten what you were trying to accomplish! Putting comments in your code is a way of making notes and passing them to people with whom you are working and/or your future self. By using the `#` in your code, you can put comments in your code that R will not execute but will tell you what is going on.  

```{r, eval = FALSE}
# R will not read this line of code when I execute the script
data(mtcars) # R doesn't read this either.  This is a good place to put comments
mean(mtcars$mpg) # especially early as you are learning R
mean(mtcars$qsec) # extensive commenting of your code is invaluable for remembering what you are doing
t.test(data = mtcars, mtcars$mpg ~ mtcars$am) # so when you look at a script later, you know what you did,
t.test(data = mtcars, mtcars$qsec ~ mtcars$am) # and why you did it.

```
A more serious version of commenting on the above code might look like:

```{r, eval = FALSE}
# Estimate whether or not having an automatic vs manual transmission effects mpg or quarter second time. Note that for the variable am, automatic = 0 and manual = 1
data(mtcars) 
mean(mtcars$mpg) 
mean(mtcars$qsec) 
t.test(data = mtcars,mpg ~ am) 
t.test(mtcars$qsec ~ mtcars$am) 


```

It is very useful to comment your scripts throughout, and also to put a comment at the top of your script to give a general description of what the script does.  For example, if I were doing a bunch of analysis on the mtcars dataset I used in the previous example, my first couple lines of code might be:

```{r}
# This is code to look at the effect of manual vs automatic transmissions in cars from the mtcars data set.
# Code developed by Matt Dobra
```

### Load required libraries early

Base R is powerful, but what makes R the state-of-the-art programming language is the user written packages.

Remember, you only need to install a package once.  This is done with the `install.packages()` function.  For example:

```{r, eval = FALSE}
install.packages("tidyverse")
```

Will install the `tidyverse` package onto your computer.  Once it is installed, you need to load it into your session whenever you use it.  When scripting, it is a good idea to load in any packages early in your script.  For example, nearly every script I write includes the following line toward the very top of the script:

```{r}
library(tidyverse)
```

If I wind up needing more packages as I develop the script, I try to put their library call at the beginning of the script, not at the end.  For example, in my R scripts that I use to write my Principles of Macroeconomics course notes, I have the following code in lines 13-23, as I wind up using these packages in nearly every script.

```{r, eval = FALSE}
library(knitr) # markdown language
library(tidyverse) # Keep things tidy
library(kableExtra) # Table Formattting
library(ggthemes) # ggplot addon
library(ggpubr) # ggplot add on
library(DiagrammeR) # Makes Flowcharts
library(WDI) # World Development Indicators
library(quantmod) # Gets FRED data
library(scales) # labeling improvement for ggplots.

```

### Break your code into sections

A typical script might have 3 distinct sections

* Preamble - a section that overviews the code, loads libraries and data, etc.
* Data Wrangling - a section the transforms the loaded data into what you will be analyzing
* Statistical Analysis and Visualiztion - a section that does the econometrics and makes graphs.

Use # comments to break these up. Make the section breaks stand out and obvious. For example:

```{r}
#----------------#
# Data Wrangling #
#----------------#

################################
#~~~~~Statistical Analysis~~~~~#
################################
```

You will probably discover more useful tips as you go along.  

## Types of script

Let's talk about 2 script types here, basic scripts and **R Markdown** documents.  An R script is used to create (and recreate!) a series of commands in R.  An **R Markdown** document incorporates script elements into a high quality document that can be shared in a variety of formats.

### Basic scripts

To create a new script, use the button on the top left of RStudio.  This can also be done in the file menu or with Ctrl+Shift+N. You can run the whole script by using the run button (top right of the script pane).  The RStudio script editor is basically a fancy text editor.  This means you can't use the Enter button to run a line of code...to run a line of code, you need to put your cursor on that line and use Ctrl-Enter.  You can also highlight a line or a set of lines with your cursor and execute that set of lines with hitting Ctrl+Enter.

### R Markdown documents

You can also embed R code into a R Markdown document to make high quality documents in a variety of types:

* HTML
* LaTeX/PDF
* Word
* Slide Presentations
* Dashboards
* e-Books

There is lots of fancy formatting that goes into all of these types that we will not be going into.   Rather, we will learn how to use a relatively simple form of document called a notebook.  In fact, this document is a notebook, and you should submit assignments as R Markdown Notebooks.  You will need to install the `rmarkdown` package, and installing `knitr` probably won't hurt either.

```{r, eval = FALSE}
install.packages("rmarkdown")
install.packages("knitr")
```

Some of the formatting code (especially the $\LaTeX$ stuff) in my notebooks is beyond the scope of what you will need; for the purposes of this class, you should be fine with very basic R Markdown formatting syntax: go to <http:\\rmarkdown.rstudio.com/lesson-1.html>, and read chapters 1-3.2 of ***R Markdown: The Definitive Guide*** by Yihui Xie <https://bookdown.org/yihui/rmarkdown/> to learn the basic (Pandoc) formatting syntax. 

Working with Notebooks is fairly simple.  Create a new notebook with the top left button.  Change the title in the preamble (or modify it using YAML options--learn about these in the Xie book if you want).  Use basic Pandoc formatting to do stuff like headers, font formatting, etc. Create **code chunks** by clicking the **Insert Chunk** button on the toolbar or by pressing **Ctrl+Alt+I**, and write your R script in the code chunks.  It is also worth learning a couple useful options for your code chunks, here are 6 I suspect you might find useful:

* `eval = FALSE` - Display the code in the finished document but don't execute it. 
* `warning = FALSE` - If R spits out a warning, don't display the warning in the finished document.  Very useful with `stargazer()`.
* `echo = FALSE` - Execute the code and display the results, but don't put the code into the final document. Very useful if you are trying to write professional looking documents.
* `include = FALSE` - execute the code but don't display the code or the results.  Very useful if you are trying to look professional looking documents.
* `message = FALSE` - Don't display any messages R tells you as a result of running code.  Very useful with `ggplot()` when specifying `geom_smooth(method = lm)`.  Very useful in conjunction with the `library()` command as well.
* `results = 'asis'` - Useful in cases where the R output is already in html format (or $\LaTeX$ format when making PDFs).

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).  The preview shows you a rendered HTML copy of the contents of the editor. *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed. 

# R Basics

## Assigning objects and basic math

R is an **object oriented** language.  We need to learn how to create and define objects. The preferred way to do so is with an arrow `<-`.  You can also use an equals sign (`=`) but that's generally frowned upon.

```{r}
q <- 42
```

This creates an object named q and q has a value of 42.

If you want to see what q is, you can either:
```{r}
print(q)
```
or simply:
```{r}
q
```
Something you will probably struggle a lot with early on is coping with the fact that R takes being a **case sensitive** language VERY, VERY seriously!
```{r, error = TRUE}
Q
```
objects can have their values overwritten:
```{r}
q <- 24
```
```{r}
q
```
Let's assign a bunch of numbers to objects:
```{r}
z <- 92
y <- 23
x <- 56
```
All the objects in memory are in the environment pane (top right).

You can remove an object with `rm()`:
```{r}
rm(q)
```
You can remove all objects with this tricky but useful command:
```{r}
rm(list=ls())
```

Object names can include numbers, periods, or underscores, but must begin with a letter.
```{r}
x.1 <- 44
x_1 <- 55
```
Any object you create should show up in your environment window. Let's clear our environment again:
```{r}
rm(list=ls())
```
We are going to need those numbers later:
```{r}
z <- 92
y <- 23
```
Objects don't have to be be just numbers.  They can be words too.
```{r}
a <- "Hello"
a
```
Even though 1 is a number, wrapping it in parenthesis means R treats it like a word, not a number>
```{r}
b <- "1"
```
We can use the `class()` command to see what type of object something is:
```{r}
class(a)
class(b)
class(y)
```
While very unexciting, we can use R as a calculator.
```{r}
2+2
24-18
45*8
84/4
2^8
abs(-42)
```
We can perform arithmetic with our variables from earlier as well:
```{r}
z+y
z-y
z*y
z/y
sqrt(y)
log(z)
exp(y)
```
We can mix and match numbers with variables
```{r}
z+4
y^2
z/y+3
```
Why doesn't this work then?
```{r, error = TRUE}
z+b
```
R reads b like a word, not a number.  Watch this though!
```{r}
z+as.numeric(b)
```
Note that b is still "1", but using `as.numeric(b)` told R to, one time only, treat b as though it were a number if possible.  Note that this won't work with a:
```{r, error = TRUE}
z+as.numeric(a)
```
R is clearly displeased with us.

We can assign objects with math too.
```{r}
q <- z/y
q
```

One last useful tip for dealing with R.  If you are using the console, you can Use the up arrow on your keyboard to scroll through your previous commands.  

## Vectors 

We can use the **concatenate** command `c()` to create a vector.
```{r}
num1 <- c(1,4,9,16,25)
num2 <- c(1,3,6,10,15)
```
vectors can also include characters:
```{r}
countries <- c("USA", "Canada", "Mexico")
```
Mathematical operations can be performed on vectors.
```{r}
num1-1
num1*3
sqrt(num1)
```
If vectors are the same length, arithmetic on vectors does **pairwise arithmetic**:
```{r}

num1+num2
num1*num2
```
We can extract elements from a vector using brackets.  
```{r}
num1[2] 
num2[4]
```
We can extract all but certain elements with the negative sign:
```{r}
num1[-3]
```
we can also extract based on a condition:
```{r}
num2[num2>5]
```
## Manipulating data

### Importing data
Many data sets are built into R.  To see what data sets are readily available, type:
```{r}
data()
```
Let's say we want to play with the iris data (people love that one, I don't know why):

```{r}
data(iris)
```
Now the iris data is loaded into the environment.  

There are also many packages that are full of data, especially in support of various textbooks. Here are a few we will make extensive use of throughout the semester: 
```{r, eval = FALSE}
install.packages("wooldridge")
install.packages("AER")
install.packages("fivethirtyeight")
```
```{r}
library(wooldridge)
library(AER)
library(fivethirtyeight)
```

These lines installed three sets of data -- two that accompany popular econometrics textbooks, and one with a repository of data from the **fivethirtyeight** website.  Now, a `data()` call will list a lot more available databases!

Let's load in the CPS1985 dataset from the AER package:
```{r}
data(CPS1985)
```
Let's put this aside for now and we will get back to it shortly.

It is not uncommon to find data that is in .csv (Comma Separated Values) format.  This is pretty easy to import into R using the `read.csv()` function.  If you find data in .xlsx (Microsoft Excel) format, just save your spreadsheet as a .csv file and this method works too.

Let's do this using the dataset they make use of in the Wehde textbook.  In section 4.4 of the book, they give a link to <http://crcm.ou.edu/epscordata/>; here you will find bunch of survey data from the state of Oklahoma. Enter your email address in the "Request Data" box, and you should soon get an email that will allow you do download a few .csv files.  Download the one that looks something like this **w1_w22_longdata.csv** and save it to your computer desktop (or somewhere else easy to find).

This file is HUGE and opening it using Excel might crash your system.  It should be a breeze for R though.  How do you open it in R?  The easiest way is to incorporate the `file.choose()` option in the `read.csv()` function.  This will open a dialogue box and you can just go find the file.  This code also then puts that data set into an object called data1:

```{r, eval = FALSE}
data1 <- read.csv(file.choose(), header = TRUE)

```

A better way to do this is to just put the whole path in the `read.csv()` command.  
```{r}
data1 <- read.csv("C:/Users/mattd/Dropbox/Courses I Teach/Applied Econometrics/Data/w1_w22_longdata.csv", header = TRUE)
```

Obviously, you want to find the file path on your own computer!  

Once we manipulate this later, we will see how to save data using the `write.csv` command.

There are many other utilities for importing data into R, including packages such as `readxl`, `haven`, `Hmisc`, and `foreign`.  If the data exists in a somewhat common format, somebody has written the package to import it into R!

### Inspecting the data

Now that this dataset is an object in memory (data1), you could just click on data1 in the environment window and look at the spreadsheet. 

To see how big the data is, try the commands `dim()`, `ncol()`, and `nrow()`

```{r}
dim(data1)
ncol(data1)
nrow(data1)
```
The `dim` command tells us the *dimensions* of the data set; there are `r nrow(data1)` rows/observations and `r ncol(data1)` columns/variables.  

The Wehde textbook only looks at the Wave 12 (Fall 2016) data, so let's filter that out using the `dplyr` tools we saw in a previous lesson.

```{r}
data1 <- data1 %>% # This line says that we are going to overwrite object data1
    filter(wave_id == "Wave 12 (Fall 2016)") # with everything currently in object data1 where wave_id says "Wave 12 (Fall 2016)" 
```
Now if we look at the dimensions of data1, we see
```{r}
dim(data1)
```
Now, if you click on the data1 object and inspect it, you will see a lot of NA values--basically, lots of survey questions were not asked in this wave.  Let's get rid of our NA columns.  This code is a little tricky.
```{r}
data1 <- data1[ , !apply(is.na(data1), 2, all)] 
```
You may want to keep this data set for later (you can use it as you read through the Wehde et. al. text to replicate their work), so let's save it as a csv file called Oklahoma.csv using the `write.csv()` command.  

```{r}
write.csv(data1, "C:/Users/mattd/Dropbox/Courses I Teach/Applied Econometrics/Data/Oklahoma.csv", row.names = FALSE)
```
And now, if I ever want to use this data, I can write:
```{r}
data2 <- read.csv("C:/Users/mattd/Dropbox/Courses I Teach/Applied Econometrics/Data/Oklahoma.csv", header = TRUE)
```
And I get my file back!

Let's return to exploring data and use the CPS1985 data we loaded earlier.  If you don't have it in your environment:
```{r}
library(AER)
data(CPS1985)
```
For most inbuilt datasets, you can get information about the dataset using the help command.
```{r}
?CPS1985
```
Let's see the dimensions of this dataset:
```{r}
dim(CPS1985)
```
So this is 534 rows and 11 columns.  IMPOTANT: R always does rows first, columns second.  Remembering this will help!

We can also learn about about what the data generally look like by using the `head()` function:
```{r}
head(CPS1985)
```
We can also use square brackets to subset bits of data.  Remember, brackets use the rows, columns convention mentinoed above.  Here I look at rows 222-225.  I leave the column part empty so I get all of the columns:

```{r}
CPS1985[222:225,]
```
We can also do logical operators here.  Let's look at every row where age is over 60:

```{r}
CPS1985[CPS1985$age>60 , ]
```
What about if we only want to see unmarried white females in the south?  Here, we need to use the double == sign (== is the boolean operator for "is equal to"), put quotes around the stuff that aren't numbers, and incorporate a bunch of ampersands:
```{r}
CPS1985[CPS1985$ethnicity == "cauc" & CPS1985$region == "south" & CPS1985$gender == "female" & CPS1985$married == "no", ]
```
Typing all that CPS1985$ stuff gets annoying.  Understanding dollar sign notation is essential, but to make our lives a bit easier, this might be a good place to `attach()` our data:
```{r}
attach(CPS1985)
```
And now that previous command can be simplified as:
```{r}
CPS1985[ethnicity == "cauc" & region == "south" & gender == "female" & married == "no", ]
```
Sometimes it is easier to filter out everything BUT a certain group.  This is accomplished with not equals signs (!=) and/or negative signs.  This next line shows me the subset of data that is males over the age of 55 who are not in management.  Because this subgroup only includes males, the gender column is irrelevant, so I'm getting rid of the 7th column.  
```{r}
CPS1985[gender == "male" & occupation != "management" & age>55, -7]
```
Keep in mind, we can get to the same place using `dplyr`, but learning how to subset is still useful. 

```{r}
CPS1985 %>% 
    filter(gender == "male") %>% 
    filter(occupation != "management") %>% 
    filter(age>55) %>% 
    select(-gender)
```
Sometimes subsetting is the better approach, sometimes dplyr is better, it just depends on what you are trying to accomplish.  Both have their place.  As you learn more about using R, you will find that there are usually a lot of ways to accomplish the same thing.
 
Another useful way to look at the data is via the `str()` function
```{r}
str(CPS1985)
```
This tells us about the types of variables we have in our data frame.  The first 4 are numeric, the rest are factor (categorical) variables.  

You can also see what the variable type is with the `class()` command.

```{r}
class(wage)
class(age)
class(married)
class(union)
```
What went wrong with the `class(union)` command?  It turns out there is a function in base R called `union`, so R is not sure what you are referring to.  This is a case where you are stuck using \$ notation, even though we used `attach()` on the data set.  This is why our original `attach()` command spit out a weird error message too.

```{r}
class(CPS1985$union)
```
If you have a categorical/factor variable, you can use the `levels` function to see what all the possible values are:
```{r}
levels(ethnicity)
levels(occupation)
```
We can get a generic summary of the data with the `summary()` command.
```{r}
summary(CPS1985)
```
Note that R provides different output for the different types of data.  For the numeric data, R gives us quantitative summary statistics -- means, min/max, and quartiles.  For the categorical data, we get raw counts.

Sometimes, datasets will code a categorical variable as a number.  Here, I will code a variable called female which will be 1 for females and 0 for males. So as to not overwrite the CPS1985 data in our environment memory, I will first clone the CPS1985 into an object called tempdata and then create a new variable called female with the `ifelse()` command.

```{r}
tempdata <- CPS1985
tempdata$female <- ifelse(tempdata$gender == "female", 1, 0)
tempdata[200:209,c(7,12)]
```
You can see from the output that my code apparently worked. We know that the female variable is essentially categorical, but what happens when we inspect the class?
```{r}
class(tempdata$female)
```
And if we summarize it:
```{r}
summary(tempdata$female)
```
This tells us that 45.88% of our data is female, which is useful, but we might want to force R to either treat the female variable as a factor, or convert it to a factor.
```{r}
summary(as.factor(tempdata$female))
```
Now it is giving counts for us. 
We could recode the variable as a factor:
```{r}
tempdata$female <- factor(tempdata$female, levels = c(0,1), labels = c("male", "female"))
summary(tempdata$female)
```
Another useful trick is to combine square brackets with our functions.  Let's look at the `summary` of wage data, split out by male vs female.
```{r}
summary(wage[gender == "male"])
summary(wage[gender == "female"])
```
As a slightly more advanced method of calculating these means, you might want to learn more about the `apply` family of functions.  This line of code uses `tapply` to calculate the mean wage by gender in the dataset.  

```{r}
tapply(X = wage, INDEX = gender, FUN = mean)
```
Though personally, I would probably stick with `dplyr` for this:
```{r}
CPS1985 %>% 
    group_by(gender) %>% 
    summarize(wage = mean(wage))
```

Maybe we want to make a new variable in our dataset to highlight which of our individuals are unmarried females in management.  The `ifelse` command works, but another useful method is to use the `cbind` command.  First, let's create a vector called femanager:

```{r}
femanager <- gender == "female" & occupation == "management"
summary(femanager)
```
You can see that this is an object with 533 observations that are TRUE or FALSE.  So apparently we have 21 female managers in our data.  We can use *cbind* to add it to our data:

```{r}
tempdata <- cbind(CPS1985, femanager)
class(tempdata$femanager)
```
The class is "logical", but this can be converted to factor easily:
```{r}
tempdata$femanager <- factor(tempdata$femanager, levels = c(FALSE,TRUE), labels = c("no", "yes"))
class(tempdata$femanager)
summary(tempdata$femanager)
```

# Conclusion

This concludes our introduction to coding in R.  The focus has been on what we might need for this class; there is a lot more out there if you want to develop a deeper understanding of R.  Next, we will turn our attention to using R to review what you (hopefully!) learned in your introduction to statistics class.  

