---
title: "Intro to Regression"
output: html_notebook
---

This notebook contains code from Chapter 5 of [Prelude to Econometrics Using R](https://bookdown.org/mattdobra/Prelude/).  

The companion site to [Prelude to Econometrics Using R](https://bookdown.org/mattdobra/Prelude/) can be found at [https://mattdobra.github.io/EconometricsLectureNotes/](https://mattdobra.github.io/EconometricsLectureNotes/).

Libraries needed for this notebook:

```{r}
library(tidyverse)
library(wooldridge)
library(AER)
library(stargazer)
library(quantmod)
library(tidyquant)
```

Preload and attach data:

```{r}
data(gpa2)
data(CPS1985)
attach(CPS1985)
data(vote1)
data(k401k)
data(StrikeDuration)
data(gpa2)
data(meap01)
data(wine)
data(airfare)

```

# Chapter 5

## Prelude to Regression: The Correlation Test

```{r echo = TRUE}
cor(wage, education)
```

```{r, echo = TRUE, message = FALSE}
CPS1985 %>% ggplot(aes(x = education, y = wage)) +
    geom_point() +
    geom_smooth(method = lm)
```

```{r echo = TRUE}
cor.test(education, wage)
```

```{r, message = FALSE, echo = TRUE}
CPS1985 %>% ggplot(aes(x = experience, y = wage)) +
    geom_point() +
    geom_smooth(method = lm)
```

```{r echo=TRUE}
cor.test(wage, experience)
```

```{r echo = TRUE}
CPS1985 %>% ggplot(aes(x = wage)) +
    geom_histogram(binwidth = 2)
```

```{r echo = TRUE}
cor.test(wage, experience, method = "spearman", exact = FALSE)
```

## Ordinary Least Squares

\begin{equation}
Y_{i} = \hat{\alpha} + \hat{\beta} X_{i}+ \epsilon_{i} 
\end{equation}

## Estimating an OLS

```{r message = FALSE}
CPS1985 %>% ggplot(aes(x = education, y = wage)) +
    geom_point() 
```

```{r, echo = TRUE}
lm(wage ~ education, data = CPS1985)
```

## Obtaining Useful Results

```{r, echo = TRUE, warning = FALSE}
reg1 <- lm(wage ~ education, data = CPS1985)
```


```{r, echo = TRUE}
attributes(reg1)
```

```{r, echo = TRUE}
summary(reg1)
```

```{r, echo = TRUE, warning = FALSE}
stargazer(reg1, type = "text")
```

## Predicted Values and Residuals

```{r, echo = TRUE}
resids <- reg1$residuals
preds <- reg1$fitted.values
tempdata <- cbind(CPS1985, preds, resids)
tempdata %>% 
    select(c(wage,education, preds, resids)) %>% 
    slice(1:10)
```

## Look at the stars, look how they shine for you

```{r, warning = FALSE, echo = TRUE}
stargazer(reg1, type = "text")
```

```{r echo = TRUE}
cor(wage, education)
cor(wage, education)^2
```

## Visualizing the Regression Line

```{r, message = FALSE}
CPS1985 %>% ggplot(aes(x = education, y = wage)) +
    geom_point() +
    geom_smooth(method = lm,
                se = FALSE)
```

## Introducing Data Transformations

```{r, message = FALSE, echo = TRUE}
airfare %>% ggplot(aes(x = dist, y = fare)) +
    geom_point(size = .7) + 
    geom_smooth(method=lm)
```

```{r echo = TRUE}
reg2 <- lm(fare ~ dist, data = airfare)
```

```{r, warning = FALSE, echo = TRUE}
stargazer(reg2, type = "text")
```

```{r echo= TRUE}
airfaretemp <- airfare %>% 
    select(fare, dist) %>% 
    mutate(dist100 = dist/100)
airfaretemp[c(1,5,9,13,17),]
```

```{r, echo = TRUE, warning = FALSE}
reg2a <- lm(fare ~ dist100, data = airfaretemp)
stargazer(reg2, reg2a, type = "text")
```

## Does Alpha Matter?

```{r, echo = TRUE, message = FALSE}
gpa2 %>%  ggplot(aes(x = sat, y = colgpa)) +
    geom_point(size = .7) +
    geom_smooth(method = lm) +
  labs(y = "GPA",
       x = "SAT Score") +
  theme_classic()
  
```

```{r, echo = TRUE, warning = FALSE}
reg3 <- lm(colgpa ~ sat, data = gpa2)
stargazer(reg3, type = "text")
```

## Alternative Hypothesis Tests

```{r, echo = TRUE, warning = FALSE, message = FALSE}
firstsolar <- tq_get('FSLR', # firstsolar Ticker
               from = "2017-01-01",
               to = "2018-01-01",
               get = "stock.prices")
vanguard500 <- tq_get('VFINX',  # vanguard500 Ticker
               from = "2017-01-01",
               to = "2018-01-01",
               get = "stock.prices")
coke <- tq_get('KO',  # Coke Ticker
               from = "2017-01-01",
               to = "2018-01-01",
               get = "stock.prices")
spx <- tq_get('^GSPC',  # S&P 500 Ticker
               from = "2017-01-01",
               to = "2018-01-01",
               get = "stock.prices")
tbill <- tq_get('^IRX', # 90 day T-bill
               from = "2017-01-01",
               to = "2018-01-01",
               get = "stock.prices")
# Convert stocks prices into daily rates of return
firstsolar <- firstsolar %>% 
    select(date, adjusted) %>% 
    mutate(lnadjusted = log(adjusted)) %>% 
    mutate(firstsolar = lnadjusted - lag(lnadjusted)) %>% 
    select(date, firstsolar)
vanguard500 <- vanguard500 %>% 
    select(date, adjusted) %>% 
    mutate(lnadjusted = log(adjusted)) %>% 
    mutate(vanguard500 = lnadjusted - lag(lnadjusted)) %>% 
    select(date, vanguard500)
coke <- coke %>% 
    select(date, adjusted) %>% 
    mutate(lnadjusted = log(adjusted)) %>% 
    mutate(coke = lnadjusted - lag(lnadjusted)) %>% 
    select(date, coke)
spx <- spx %>% 
    select(date, adjusted) %>% 
    mutate(lnadjusted = log(adjusted)) %>% 
    mutate(spx = lnadjusted - lag(lnadjusted)) %>% 
    select(date, spx)
# t bill quotes are annualized, convert to daily
tbill <- tbill %>% 
    select(date, adjusted) %>% 
    mutate(tbill = adjusted/360) %>% 
    select(date, tbill)
# Combine all the datasets
returns <- spx 
returns <- merge(returns, vanguard500, by = "date")
returns <- merge(returns, firstsolar, by = "date")
returns <- merge(returns, coke, by = "date")
returns <- merge(returns, tbill, by = "date")
returns <- drop_na(returns)
```

```{r, warning = FALSE, echo = TRUE}
capm1 <- lm(coke-tbill~spx-tbill, data = returns)
capm2 <- lm(firstsolar-tbill~spx-tbill, data = returns)
capm3 <- lm(vanguard500 - tbill ~ spx - tbill, data = returns)
stargazer(capm1, capm2, capm3, type = "text")
```

```{r echo = TRUE}
confint(capm1) # Coke Model
```

```{r echo = TRUE}
confint(capm2) # First Solar Model
confint(capm3) # Vanguard Model
```

## Not All Regressions Have Significant Results

```{r, message = FALSE, echo = TRUE}
meap01 %>% ggplot(aes(x = exppp, y = math4)) +
    geom_point(color = "darkslateblue",
               alpha = .6) +
    geom_smooth(method = lm)
```


```{r, warning = FALSE, echo = TRUE}
reg4 <- lm(math4 ~ exppp, data = meap01)
stargazer(reg4, type  = "text")
```

## Signs, Signs, Everywhere There's Signs

```{r, message = FALSE, echo = TRUE}
vote1 %>% ggplot(aes(x = shareA, y = voteA)) +
    geom_point(color="indianred") +
    geom_smooth(color = "dodgerblue", method = lm, se = FALSE) +
  labs(y = "Vote Share",
       x = "Campaign Spending") +
  theme_classic()
```

```{r, warning = FALSE, echo = TRUE}
reg5 <- lm(voteA ~ shareA, data = vote1)
stargazer(reg5, type = "text")
```

## Multivariate Regression

```{r, echo = TRUE, warning = FALSE}
reg1 <- lm(wage ~ education, data = CPS1985)
reg6 <- lm(wage ~ age, data = CPS1985)
reg7 <- lm(wage ~ education + age, data = CPS1985)
stargazer(reg1, reg6, reg7, type = "text")
```

## Predictive modeling

```{r, echo = TRUE, warning = FALSE}
reg6 <- lm(wage ~ age, data = CPS1985)
reg7 <- lm(wage ~ education + age, data = CPS1985)
stargazer(reg6, reg7, type = "text")
```


```{r echo = TRUE}
a <- data.frame(Model = "Bivariate", Residual = resid(reg6))
b <- data.frame(Model = "Multivariate", Residual = resid(reg7))
residdata <- rbind(a,b)
ggplot(residdata) +
  geom_boxplot(aes(y = Residual, fill = Model))
```


```{r echo = TRUE}
sd(resid(reg6))
sd(resid(reg7))
```

## Multivariate Models, Part Deux

```{r, warning = FALSE}
stargazer(reg1, reg6, reg7, type = "text")
```

## The Importance of Controls

```{r, echo = TRUE, warning = FALSE}
stargazer(reg4, type = "text")
```

```{r, echo = TRUE, warning = FALSE}
reg4z <- lm(math4 ~ exppp + lunch, data = meap01)
stargazer(reg4, reg4z, type = "text")
```

### Statistical vs. Economic Significance

```{r, echo = FALSE}
stargazer(reg4, reg4z, type = "text")
```

```{r, echo = TRUE}
meap01 %>% 
  select(exppp, math4) %>% 
  stargazer(data = ., type = "text")
```

```{r, echo = TRUE}
meap01 %>% 
  ggplot(aes(x = math4)) +
  geom_histogram(binwidth = 2, color = "black") +
  scale_x_continuous(expand=c(0,0)) +
  scale_y_continuous(expand=c(0,0)) +
  theme_classic()
```

```{r, echo = FALSE, warning=FALSE, message = FALSE}
detach(CPS1985)
```