---
title: "Regression Assumptions"
output: html_notebook
---

This notebook contains code from Chapter 6 of [Prelude to Econometrics Using R](https://bookdown.org/mattdobra/Prelude/).  

The companion site to [Prelude to Econometrics Using R](https://bookdown.org/mattdobra/Prelude/) can be found at [https://mattdobra.github.io/EconometricsLectureNotes/](https://mattdobra.github.io/EconometricsLectureNotes/).

Libraries needed for this notebook:

```{r}
library(tidyverse)
library(wooldridge)
library(AER)
library(stargazer)
library(sandwich)
library(lmtest)
```

Preload data:

```{r}
data(wage1)
data(CPS1985)
data(ceosal1)
data(nyse)
data(smoke)
data(vote1)
data(hprice3)
data(infmrt)

```

# Chapter 6

## Assumption 1: The Linear Regression Model is "Linear in Parameters"  

```{r, echo = TRUE, message = FALSE}
ceosal1 %>% ggplot(aes(y = salary, x = sales)) + 
    geom_point() +
    theme_classic() +
    geom_smooth(method=lm)
```


```{r, echo=TRUE, warning = FALSE}
reg1a <- lm(salary ~ sales, data = ceosal1)
stargazer(reg1a, type = "text")
```

```{r echo = TRUE}
tempdata <- ceosal1 %>% 
    mutate(lnsalary = log(salary)) %>% 
    mutate(lnsales = log(sales))
```

```{r, echo = TRUE, warning = FALSE}
tempdata %>% ggplot(aes(x = lnsales, y = lnsalary)) +
    geom_point() +
    theme_classic() +
  geom_smooth(method = lm)
    
```


```{r, warning = FALSE, echo = TRUE}
reg1b <- lm(lnsalary ~ lnsales, data = tempdata)
stargazer(reg1b, type = "text")
```


```{r, echo = TRUE, warning = FALSE}
stargazer(reg1a, reg1b, type = "text")
```

```{r, warning = FALSE, echo = TRUE}
reg1c <- lm(lnsalary ~ sales, data = tempdata)
reg1d <- lm(salary ~ lnsales, data = tempdata)
stargazer(reg1a, reg1d, reg1c, reg1b, type = "text")
```


```{r echo = TRUE}
hprice3 %>% ggplot(aes(x = age, y = price)) +
    geom_point(color = "cornflowerblue") +
    theme_classic() +
    labs(title = "House Value and Age") 
```

```{r, warning = FALSE, echo = TRUE}
reg1e <- lm(price ~ age, data = hprice3)
reg1f <- lm(price ~ age + I(age^2), data = hprice3)
reg1g <- lm(price ~ age + agesq, data = hprice3)
stargazer(reg1e, reg1f, reg1g, type = "text")
```

```{r, message = FALSE, echo = TRUE}
hprice3 %>% ggplot(aes(x = age, y = price)) +
    geom_point(color = "cornflowerblue",
               alpha = .4) +
    theme_classic() +
    labs(title = "House Value and Age",
         y = "Price",
         x = "Age") +
    geom_smooth(method = lm, 
                color = "palegreen4",
                se = FALSE) +
    geom_smooth(method = lm, 
                formula = y ~ x + I(x^2),
                color = "indianred3",
                se = FALSE) +
    scale_y_continuous(labels=scales::dollar_format())
```

## Assumption 2: The Average of the Error Term is 0


```{r, echo = TRUE}
summary(reg1b)
```


```{r, echo = TRUE}
tempdata$resid <- reg1b$residuals
head(tempdata[c(1,3,13,14,15)],)
```


```{r, echo = TRUE}
mean(tempdata$resid)
```

```{r, echo = TRUE}
format(mean(tempdata$resid), scientific = FALSE)
```

## Assumption 3: The Independent Variable is Uncorrelated with the Error Term

```{r, echo = TRUE}
cor(tempdata$lnsales, tempdata$resid)
```

## Assumption 4: The Error Term is not Serially Correlated.

```{r echo = TRUE}
tempdata2 <- nyse 
tempdata2 %>% ggplot(aes(x = t, y = price)) +
    geom_point() +
    geom_line()
```


```{r echo = TRUE}
reg4a <- lm(price ~ t, data = tempdata2)
tempdata2$resid <- reg4a$residuals
tempdata2 %>% ggplot(aes(x = t, y = resid)) +
    geom_point() +
    geom_line()
```

## Assumption 5: Homoskedasticity of the Error Term

```{r echo = TRUE}
reg5a <- lm(voteA ~ shareA, data = vote1)
plot(reg5a$residuals ~ reg5a$fitted.values)
abline(a = 0, b = 0)
```

```{r, echo = TRUE}
reg5b <- lm(cigs ~ income, data = smoke)
plot(reg5b$residuals ~ reg5b$fitted.values)
abline(a = 0, b = 0)
```


```{r echo = TRUE}
coeftest(reg5b, vcovHC)
```

```{r, warning = FALSE, echo = TRUE}
stargazer(reg5b, coeftest(reg5b, vcovHC), type = "text")
```

```{r, echo = TRUE}
reg5z <- lm(infmort ~ afdcper + physic, data = infmrt)
stargazer(reg5z, coeftest(reg5z, vcovHC), type = "text")
```

## Assumption 6: No Independent Variable is a Perfect Linear Function of other Explanatory Variables.

```{r, warning = FALSE, echo = TRUE}
reg6a <- lm(voteA ~ shareA + shareA, data = vote1)
stargazer(reg6a, type = "text")
```

```{r, error = TRUE, echo = TRUE}
tempdata3 <- vote1
tempdata3$shareAclone <- tempdata3$shareA
reg6b <- lm(voteA ~ shareA + shareAclone, data = tempdata3)
stargazer(reg6b, type = "text")
summary(reg6b)
```

```{r echo = TRUE}
tempdata3$six <- 6
reg6c <- lm(voteA ~ shareA + six, data = tempdata3)
summary(reg6c)
```

```{r echo = TRUE}
set.seed(8675309)
tempdata4 <- tempdata3 %>% 
    mutate(shareArand = shareA + runif(173, min = -1, max = 1))
cor(tempdata4$shareA, tempdata4$shareArand)
```


```{r, warning = FALSE, echo = TRUE}
reg6d <- lm(voteA ~ shareA + shareArand, data = tempdata4)
reg6e <- lm(voteA ~ shareArand, data = tempdata4)
stargazer(reg6a, reg6e, reg6d, type = "text")
```

## Assumption 7: Normality of Error Terms


```{r echo = TRUE}
reg7a <- lm(voteA ~ shareA, data = vote1)
qqnorm(reg7a$residuals)
qqline(reg7a$residuals)
```

```{r echo = TRUE}
qqnorm(reg1b$residuals)
qqline(reg1b$residuals)
```


```{r echo = TRUE}
plot(reg7a)
```
