---
title: "Regression Independent Variables"
output: html_notebook
---

This notebook contains code from Chapter 7 of [Prelude to Econometrics Using R](https://bookdown.org/mattdobra/Prelude/).  

The companion site to [Prelude to Econometrics Using R](https://bookdown.org/mattdobra/Prelude/) can be found at [https://mattdobra.github.io/EconometricsLectureNotes/](https://mattdobra.github.io/EconometricsLectureNotes/).

Libraries needed for this notebook:

```{r}
library(tidyverse)
library(wooldridge)
library(AER)
library(stargazer)
```

Preload and attach data:

```{r}
data(CPS1985)
data(HousePrices)
data(diamonds)
data(gpa2)
data(TeachingRatings)
data(CPS1988)
data(GrowthDJ)
data(CollegeDistance)
data(happiness)
data(beauty)
data(campus)
data(twoyear)

```

# Chapter 7

## Dummy Variables

```{r echo = TRUE}
head(CPS1985)
```


```{r echo = TRUE}
summary(CPS1985$gender)
```


```{r echo = TRUE}
class(CPS1985$gender)
```


```{r echo = TRUE}
tempdata <- CPS1985
tempdata$female <- ifelse(tempdata$gender == "female", 1, 0)
tempdata$male <- ifelse(tempdata$gender == "male", 1, 0)
```

```{r echo = TRUE}
tempdata[168:177, c(7, 12, 13)]
```

```{r echo = TRUE}
summary(CPS1985$occupation)
```

## Anything ANOVA Does, Regression Does Better

```{r echo = TRUE}
t.test(wage ~ gender, data = CPS1985)
```

```{r, echo = TRUE}
summary(aov(wage ~ gender, data = CPS1985))
```


```{r echo = TRUE}
oneway.test(wage ~ gender, data = CPS1985)
```

```{r, warning = FALSE, echo = TRUE}
reg1a <- lm(wage ~ female, data = tempdata)
stargazer(reg1a, type = "text")
```

```{r, echo = FALSE}
summary(aov(wage ~ gender, data = tempdata))
```

```{r, echo = FALSE}
anova(reg1a)
```

```{r, warning = FALSE}
stargazer(reg1a, type = "text")
```

```{r echo = TRUE}
tempdata %>% ggplot(aes(fill = gender, y = wage)) +
    geom_boxplot() +
    scale_fill_brewer(palette="Accent")
```

```{r, warning = FALSE, echo = TRUE}
reg1b <- lm(wage ~ gender, data = tempdata)
stargazer(reg1a, reg1b, type = "text")
```

```{r, warning = FALSE, echo = TRUE}
reg1c <- lm(wage ~ male, data = tempdata)
stargazer(reg1a, reg1c, type = "text")
```

```{r, warning = FALSE, echo = TRUE}
reg1d <- lm(wage ~ relevel(gender, "female"), data = tempdata)
stargazer(reg1c, reg1d, type = "text")
```

```{r, warning = FALSE, echo = TRUE}
reg1e <- lm(wage ~ male + female, data = tempdata)
stargazer(reg1e, type = "text")
```

```{r echo = TRUE}
tempdata$tempvar <- tempdata$male + tempdata$female
tempdata[168:177, c(7, 12, 13, 14)]
```

```{r, warning = FALSE, echo = TRUE}
reg1f <- lm(wage ~ female + male - 1, data = tempdata)
stargazer(reg1f, type = "text")
```

```{r, warning = FALSE, echo = TRUE}
reg2a <- lm(wage ~ occupation, data = tempdata)
stargazer(reg2a, type = "text")
```

```{r, warning = "FALSE", echo = TRUE}
reg2b <- lm(wage ~ relevel(occupation, "services"), data = tempdata)
stargazer(reg2b, type = "text")
```


```{r, warning = FALSE, echo = TRUE}
tempdata$occupation <- relevel(tempdata$occupation, "services")
reg2c <- lm(wage ~ occupation, data = tempdata)
stargazer(reg2c, type = "text")
```

```{r, warning = FALSE, echo = TRUE}
reg3aa <- lm(wage ~ gender, data = tempdata)
reg3ab <- lm(wage ~ married, data = tempdata)
reg3ac <- lm(wage ~ gender + married, data = tempdata)
reg3ad <- lm(wage ~ gender*married, data = tempdata)
stargazer(reg3aa, reg3ab, reg3ac, reg3ad, type = "text")
```

```{r, warning = FALSE, echo = TRUE}
stargazer(reg3ad, type = "text")
```

## Combining Numeric and Categorical Independent Variables

```{r, warning = FALSE, echo = TRUE}
reg4a <- lm(wage ~ education + gender, data = CPS1985)
stargazer(reg4a, type = "text")
```

```{r, echo = TRUE}
CPS1985 %>% ggplot(aes(x = education, y = wage, color = gender)) +
    geom_point() +
    scale_color_manual(values = c("cornflowerblue","indianred")) +
    geom_abline(slope = .751, intercept = .218, color = "darkblue", size = 1.2) +
    geom_abline(slope = .751, intercept = .218-2.124, color = "firebrick4", size = 1.2) +
    theme_classic()
```


```{r echo = TRUE}
head(HousePrices)
```

```{r echo = TRUE}
summary(HousePrices)
```


```{r, warning = FALSE, echo = TRUE}
reg4b <- lm(price ~ lotsize + bedrooms + bathrooms + stories + driveway + recreation + fullbase + gasheat + aircon + garage + prefer, data = HousePrices)
stargazer(reg4b, type = "text")
```

```{r echo = TRUE}
yourhouse <- data.frame(lotsize = 5000, bedrooms = 4, bathrooms = 2, stories = 2, driveway = "yes", recreation = "yes", fullbase = "no", gasheat = "no", aircon = "no", garage = 2, prefer = "yes")
```

```{r echo = TRUE}
predict(reg4b, newdata = yourhouse)
```


```{r echo = TRUE}
yourhouse[2,] <- yourhouse[1,]
```

```{r echo = TRUE}
yourhouse[2,]$bedrooms <- 5
yourhouse[2,]$bathrooms <- 3
yourhouse[2,]$fullbase <- "yes"
yourhouse[2,]$gasheat <- "yes"
yourhouse[2,]$aircon <- "yes"
```

```{r echo = TRUE}
predict(reg4b, newdata = yourhouse)
```

```{r echo = TRUE}
head(diamonds)
summary(diamonds)
```


```{r, warning = FALSE, echo = TRUE}
diamonds2 <- as.data.frame(diamonds)
diamonds2$cut <- factor(diamonds2$cut, ordered = FALSE)
diamonds2$color <- factor(diamonds2$color, ordered = FALSE)
diamonds2$clarity <- factor(diamonds2$clarity, ordered = FALSE)
diamonds2$color <- relevel(diamonds2$color, c("J")) 
reg4c <- lm(price ~ carat +I(carat^2) + cut + color + clarity, data = diamonds2)
stargazer(reg4c, type = "text")
```

```{r echo = TRUE}
mydiamond <- data.frame(carat = 32, clarity = "IF", cut = "Ideal", color = "D")
predict(reg4c, newdata = mydiamond)
```

```{r echo = TRUE}
summary(diamonds2$carat)
```

```{r, warning = FALSE, echo = TRUE}
reg4d <- lm(colgpa ~ I(sat/100) + athlete + female, data = gpa2)
stargazer(reg4d, type = "text")
```


```{r, warning = FALSE, echo = TRUE}
reg4e <- lm(colgpa ~ I(sat/100) + athlete*female, data = gpa2)
stargazer(reg4d, reg4e, type = "text")
```


```{r, warning = FALSE, echo = TRUE}
reg4f <- lm(eval~beauty + gender + minority + native + credits, data = TeachingRatings)
reg4g <- lm(eval~beauty + gender*minority + native + credits, data = TeachingRatings)
stargazer(reg4f, reg4g, type = "text")
```

## Interactions with Numeric Variables

```{r, warning = FALSE, echo = TRUE}
reg5a <- lm(log(wage) ~ experience + I(experience^2) + education*ethnicity, data = CPS1988)
stargazer(reg5a, type = "text")
```

```{r, warning = FALSE, echo = TRUE}
reg5b <- lm(math4 ~ exppp + lunch, data = meap01)
stargazer(reg5b, type = "text")
```



```{r warning = FALSE, echo = TRUE}
reg5c <- lm(math4 ~ exppp * lunch, data = meap01)
stargazer(reg5b, reg5c, type = "text")
```


