---
title: "Panel Regression"
output: html_notebook
---

# Overview

**Panel Regression** is a technique used for data that has both a *cross sectional* and a *time series* component.  In other words, data where each individual/country/company/etc. in the data set is observed at multiple points in time.  While there are panel versions of most of the models we have encountered thus far (e.g. logit/probit, Tobit, negbin, ...), we will focus only on the linear model.  

As usual, we begin with loading in some essential packages. We will need the `plm` package, so install it if you have not done so already.  


```{r, eval = FALSE}
install.packages("plm")
```

```{r, warning=FALSE, message = FALSE}
library(tidyverse)
library(stargazer)
library(plm)
library(wooldridge)
library(AER)
library(fivethirtyeight)

data(USSeatBelts)
data(Fatalities)
data(Municipalities)
data(USAirlines)
data(driving)
data(airfare)
data(murder)
data(NaturalGas)
data(HealthInsurance)
data(wagepan)

```

# Censored Models